---
title: "MPAs and climate change: analysis and plotting"
author: "Mark Ruddy"
date: 2017-04-29
  html_document:
    theme: readable
---

## Task summary

The project aims to look at projected climate warming in Marine Protected Areas. It uses climate models ([CMIP5](http://cmip-pcmdi.llnl.gov/cmip5/)) run for two emission scenarios (Representative Concentration Pathways - [RCPs](https://www.skepticalscience.com/rcp.php?t=1)) - 4.5degC and 8.5degC - developed as part of Intergovernmental Panel on Climate Change's (IPCC) Fifth Assessment Report ([AR5](http://www.ipcc-data.org/sim/gcm_monthly/AR5/)).

As part of the study, modelled temperature data need to be modified to exclude land-based temperature measures. Additionally, adequate output maps need to be created.

1. Code that removes the land surface temperatures from the CMIP5 surface temperature rasters.
2. Calculate mean temperature changes for specific geographic zones.
3. Plots of SST with land surface removed (white or black) and a pleasing SST key.


## Setup

```{r}
# Packages
library(dplyr) # Data wrangling and exploration
library(tidyr) # Data wrangling
library(ggplot2) # Plotting
library(viridis) # Perceptually sound palettes
library(gridExtra) # Grid arranged graphics
library(grid) # Grid arrangement
library(raster) # Dealing with rasters
library(rasterVis) # Dealing with rasters
library(sp) # Classes and methods for spatial data
library(ncdf4) # Required - viewing and working with climate data in netCDF format
library(rgdal) # For working with shapefiles
```

## Land surface temperature removal

### Approach


The approach I'll take is to select cells from the CMIP5 grid that fall *wholly* within the boundary of the world land surface polygon. This is because CMIP5 data is provided as a 360 \* 180 (1 \* 1 degree) grid. Raster-based analysis of temperatures - ie to quantify whole ocean or region warming - uses temperature data for each cell, so temperature values from single cells are what we're interested in. By only discarding cells that fall entirely within the land surface we ensure that SSTs for MPAs found within grid cells that overlap coastlines will be retained for analysis.

Summary of process in dummy code:


######################################
 NOT IMPLEMENTED - DUMMY CODE ###
 Mask CMIP5 with land surface ##
 "cmip" is CMIP5 netCDF read in by raster
 "land" is vector polygon land surface dataset
/> cmip.mask <- mask(x = cmip, mask = land, inverse = TRUE)

 Get MPA cells ##
 "mpas.xy" are the MPA longitude and latitude coordinate data
/> mpas <- SpatialPoints(mpas.xy) # Make MPAs into SpatialPoints class
/> mpa.cells <- mask(x = cmip, mask = mpas)

 Merge land masked CMIP5 and MPA cells ##
/> cmip.sst <- merge(cmip.mask, mpa.cells)
######################################



### Read-in data

Requires land surface, and the CMIP5 mean RCP 8.5 climate model dataset will be used here. 

Read in Marine Protected Areas
```{r}
mpas.in <- read.csv("../nn_extractedA2c.csv", header = TRUE)
mpas.xy <- mpas.in[,2:3]
rm(mpas.in)
```

Read in CMIP5 RCP8.5 mean
```{r get-cmip-test}
cmip <- "../trend_yearmean_ensemble_tos_RCP85.nc"
cmip <- raster(cmip) # Read in raster
map.extent <- c(-180, 180, -90, 90) # Reset for layer, orginal orientation was 0-360 longitude
extent(cmip) <- map.extent # Change extent

source("../revrotate.R") # Rotate raster layer to align with MPA points
cmip <- revrotate(cmip)
extent(cmip) <- map.extent

rm(map.extent, revrotate)
```

Obtain world land surface raster data

Read in [Natural Earth](http://www.naturalearthdata.com/) 1:110m world vector, then convert to simple raster.

```{r get-land-test}
# Natural Earth 110m data. Comes as zip file 
url <- "http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/110m/physical/ne_110m_land.zip" # web location

localDir <- "../data"

if (!file.exists(localDir)) {
  dir.create(localDir)
}

dwnld <- paste(localDir, basename(url), sep = "/") # zip file destination

if (!file.exists(dwnld)) {
  download.file(url = url, destfile = dwnld, method = "auto") # download from web
  unzip(dwnld, exdir = localDir) # unzip
}

land <- readOGR(dsn = localDir) # Read in shapefile using `rgdal`. No need to specify shapefile name as there's only one in the directory we created above.

rm(url, localDir, dwnld) # clean up workspace
```


```{r cmip-sst-only}
## Mask CMIP5 with land surface ##
# "cmip" is CMIP5 netCDF read in by raster
# "land" is vector polygon land surface dataset
cmip.mask <- mask(x = cmip, mask = land, inverse = TRUE)

## Get MPA cells ##
# "mpas.xy" are the MPA longitude and latitude coordinate data
mpas <- SpatialPoints(mpas.xy) # Make MPAs into SpatialPoints class

mpa.cells <- mask(x = cmip, mask = mpas)

## Merge land masked CMIP5 and MPA cells ##
cmip.sst <- merge(cmip.mask, mpa.cells)

Summary(cmip.sst)

rm(cmip.mask, mpa.cells)
```
Resulting CMIP raster has min = 0.0018 and max = 0.16 degC/year cell values.

Check for NA cells where MPAs are located

```{r}
mpa.check <- raster::extract(x = cmip.sst, y = SpatialPoints(mpas.xy)) # Use `raster::extract()` to disambiguate from `tidyr::extract()`
length(mpa.check[is.na(mpa.check)])

rm(mpa.check)
```

Should have 0 cells having temperature data intersecting with MPA locations.


## Mean temperature of geographic zones

Define latitudinal zones. We're just setting Northern hemisphere limits. Mirror image southern hemisphere limits will be derived within the function to calculate zonal means.
`raster::extract` will be used, which selects cells by their centroid from `raster::extents`. This means we won't have 

```{r define-zones}
# Find global means

# Find regional means. 
# Latitudinal zone extents

global.lims <- c(0, 90)

## Tropical
trop.lims <- c(0, 23.5)

## Sub-tropical
subtrop.lims <- c(23.5, 40) # N and S subtropical zone limits.

## Temperate
temp.lims <- c(40, 66.5) # N and S temperate zone limits.

## Polar
polar.lims <- c(66.5, 90) # N and S polar zone limits.

## All extents
zone.lims <- list(global.lims, trop.lims, subtrop.lims, temp.lims, polar.lims)
```

Make data frame to hold zone names, limits and results.
```{r zones-df}
# Initialise data frame for outputs. A tibble.
zones <- c("global", "tropical", "sub-tropical", "temperate", "polar")

zone.means <- data_frame(
  zone = zones, 
  limits = zone.lims,
  means = numeric(length = length(zones)), # Mean temperature
  count = numeric(length = length(zones)) # !is.na cell count for cross-checking data capture
)

rm(list = ls(pattern = ".*ext$|.*lims$"))
```


Calculate zonal mean temperature values.
```{r }
# Function to get raster means within extents
# raster = raster of values
# zones = df of zones, limits and empty values column for mean

source("../functions/multi.extent.means.R") # Applying raster::extract over zone dataframe

zone.means.sst <- multi.extent.means(cmip.sst, zone.means)

zone.means.sst # Sea temperatures
# zone.means.all # Sea and land temperatures

rm(zones, zone.means, multi.extent.means, extent.mean)
```

Let's check the cells captured by these limits to make sure there's no double counting. First, we'll find the number of non-NA cells in the SST raster globally, and then compare this number against the total of non-NA cells from recombining extracts made from splitting the original raster into zones. If the totals match then the latitudinal limits used develop the regional models don't overlap - cells aren't used twice. We would expect this not to be the case as `raster::extract` selects cells based on the proximity of their centroids within extents used to define sample areas.

!is.na cells in SST raster globally
```{r}
# !is.na cells in SST raster
length(getValues(cmip.sst))-summary(cmip.sst)[6]
```

Recombined !is.na cells
```{r}
sum(zone.means.sst$count[-1])
```

Cell totals from original SST raster and recombined zonal SST extracts should both match. 


## Plotting

Aim is to produce plot of CMIP raster, MPAs and land surface for four areas:

1. Global
2. The Antarctic peninsula/southern South America
3. Australia, East Indies, Polynesia
4. Europe


The world plot above contains some redundant areas near both poles. Antartica and the Arctic are warped to a large size and the poles don't contain any MPAs. We can trim these areas down to maximise the space avaiable for the rest of the globe. Something like crop at 80 degrees latitude seems appropriate. This is just for plotting, not for analysis.

Set some plot parameters for ease of use.

```{r}
# Size of axis text and guide bar text
axis_guide_text <- element_text(size = 12)
```




### Global plot
```{r}
##############
# Crop world land shapefile for better use of whole map space in plot using `raster::extract`
## Limits
crop.lim <- c(-180, 180, -82, 84) # Crop at 80 deg latitude

## Crop land
crop.land <- crop(land, extent(crop.lim))
crop.land.gg <- fortify(crop.land) # Understandable by ggplot 
##############

p.wld <- gplot(cmip) + 
  # Note1: plotting raster uses `rasterVis::gplot` here not `ggplot`.
  # Note2: plot uses original cmip dataset without land-surface temperatures removed for aesthetic reasons - would otherwise show empty NA cells at some land margins.
  geom_raster(aes(fill = value)) +
  scale_fill_viridis(direction = -1, option = "plasma", breaks = seq(0, 0.16, 0.04)) +
  geom_polygon(data = crop.land.gg, aes(x = long, y = lat, group = group), fill = "grey90") +
  geom_point(data = mpas.xy, mapping = aes(x = Centroid_Longitude, y = Centroid_Latitude), size = 0.7, colour = "black") + 
  guides(fill = guide_colorbar(title = expression(paste(degree, C,"/year")), barheight = 35, raster = TRUE)) +   
  theme(legend.title = axis_guide_text, plot.title = element_text(size = 20, hjust = 0), axis.text = axis_guide_text, legend.text = axis_guide_text) +
  labs(title = "a") +
  xlab("") +
  ylab("") +
  scale_y_continuous(limits = crop.lim[3:4] , breaks = seq(-80, 80, 20), expand = c(0,0)) +
  scale_x_continuous(breaks = seq(-180, 180, 60), expand = c(0,0)) +
  geom_hline(yintercept = c(-66.5, -40, -23.5, 23.5, 40, 66.5), linetype = 2, size = 0.3) 

ggsave(filename = "globe", plot =  p.wld, device = pdf, path = "../images", width = 13, height = 8, units = "in")

rm(lims, list = ls(patter = "^crop.*"))
```

### Antartic plot.
```{r}
##############
# Crop world land shapefile using `raster::extract`
## Limits
lims <- c(-90, -25, -80, -25) # Antarctic peninsula/southern South America

# Use `mapdata::world` land data here rather the more generalised Natural Earth .shp
world <- map_data(map = "world", interior = FALSE)

p.ant <- gplot(cmip) + 
  # Note1: plotting raster uses `rasterVis::gplot` here not `ggplot`.
  # Note2: plot uses original cmip dataset without land-surface temperatures removed for aesthetic reasons - would otherwise show empty NA cells at some land margins.
  geom_raster(aes(fill = value)) +
  scale_fill_viridis(direction = -1, option = "plasma", breaks = seq(0, 0.16, 0.04)) +
  geom_map(data = world, map = world, aes(x = long, y = lat, map_id = region), colour = "grey90", fill = "grey90") +
  geom_point(data = mpas.xy, mapping = aes(x = Centroid_Longitude, y = Centroid_Latitude), size = 0.7, colour = "black") + 
  guides(fill = guide_colorbar(title = expression(paste(degree, C,"/year")), barheight = 10, raster = TRUE)) +   
    theme(legend.title = axis_guide_text, plot.title = element_text(size = 20, hjust = 0), axis.text = axis_guide_text, legend.text = axis_guide_text) +
  labs(title = "b") +
  xlab("") +
  ylab("") +
  scale_y_continuous(limits = lims[3:4] , expand = c(-0.05, 0)) +
  scale_x_continuous(limits = lims[1:2], expand = c(-0.05,0)) +
  geom_hline(yintercept = c(-66.5, -40, -23.5, 23.5, 40, 66.5), linetype = 2, size = 0.3) 
p.ant
ggsave(filename = "ant", plot =  p.ant, device = pdf, path = "../images", width = 8, height = 8, units = "in")
rm(world, lims)
```

### Australia/East Indies plot.
```{r}
##############
# Crop world land shapefile using `raster::extract`
## Limits
lims <- c(100, 160, -50, 10) # Australia, East Indies, Polynesia

# Use `mapdata::world` land data here rather the more generalised Natural Earth .shp
world <- map_data(map = "world", interior = FALSE)

p.aus <- gplot(cmip) + 
  # Note1: plotting raster uses `rasterVis::gplot` here not `ggplot`.
  # Note2: plot uses original cmip dataset without land-surface temperatures removed for aesthetic reasons - would otherwise show empty NA cells at some land margins.
  geom_raster(aes(fill = value)) +
  scale_fill_viridis(direction = -1, option = "plasma", breaks = seq(0, 0.16, 0.04)) +
  geom_map(data = world, map = world, aes(x = long, y = lat, map_id = region), colour = "grey90", fill = "grey90") +
  geom_point(data = mpas.xy, mapping = aes(x = Centroid_Longitude, y = Centroid_Latitude), size = 0.7, colour = "black") + 
  guides(fill = guide_colorbar(title = expression(paste(degree, C,"/year")), barheight = 10, raster = TRUE)) +   
    theme(legend.title = axis_guide_text, plot.title = element_text(size = 20, hjust = 0), axis.text = axis_guide_text, legend.text = axis_guide_text) +
  labs(title = "c") +
  xlab("") +
  ylab("") +
  scale_y_continuous(limits = lims[3:4] , expand = c(-0.05, 0)) +
  scale_x_continuous(limits = lims[1:2], expand = c(-0.05,0)) +
  geom_hline(yintercept = c(-66.5, -40, -23.5, 23.5, 40, 66.5), linetype = 2, size = 0.3) 
p.aus
ggsave(filename = "aus", plot =  p.aus, device = pdf, path = "../images", width = 8, height = 8, units = "in")
rm(world, lims)
```


### Europe plot.
```{r}
##############
# Crop world land shapefile using `raster::extract`
## Limits
lims <- c(-27, 30, 30, 80) # Europe

# Use `mapdata::world` land data here rather the more generalised Natural Earth .shp
world <- map_data(map = "world", interior = FALSE)

p.eur <- gplot(cmip) + 
  # Note1: plotting raster uses `rasterVis::gplot` here not `ggplot`.
  # Note2: plot uses original cmip dataset without land-surface temperatures removed for aesthetic reasons - would otherwise show empty NA cells at some land margins.
  geom_raster(aes(fill = value)) +
  scale_fill_viridis(direction = -1, option = "plasma", breaks = seq(0, 0.16, 0.04)) +
  geom_map(data = world, map = world, aes(x = long, y = lat, map_id = region), colour = "grey90", fill = "grey90") +
  geom_point(data = mpas.xy, mapping = aes(x = Centroid_Longitude, y = Centroid_Latitude), size = 0.7, colour = "black") + 
  guides(fill = guide_colorbar(title = expression(paste(degree, C,"/year")), barheight = 10, raster = TRUE)) +   
    theme(legend.title = axis_guide_text, plot.title = element_text(size = 20, hjust = 0), axis.text = axis_guide_text, legend.text = axis_guide_text) +
  labs(title = "d") +
  xlab("") +
  ylab("") +
  scale_y_continuous(limits = lims[3:4] , expand = c(-0.02, 0)) +
  scale_x_continuous(limits = lims[1:2], expand = c(-0.03,0)) +
  geom_hline(yintercept = c(-66.5, -40, -23.5, 23.5, 40, 66.5), linetype = 2, size = 0.3) 
p.eur
ggsave(filename = "eur", plot =  p.eur, device = pdf, path = "../images", width = 8, height = 8, units = "in")
rm(world, lims)
```


### Grid plot

`gridExtra::grid.arrange` gives us the arrangement we want; with 1:4 plots on the left and a single guide on the right.

Plot with ggplot grobs (graphical objects)
```{r}
# List of plots
plist <- list(p.wld, p.ant, p.aus, p.eur)

# Get legend as grob
g <- ggplotGrob(p.wld + theme(legend.position = "right"))$grobs 
legend <- g[[which(sapply(g, function(x) x$name) == "guide-box")]]

# Remove legends from plots
plist <- lapply(plist, function(x) x + theme(legend.position="none"))

# Arrange plots in grid
mp <- grid.arrange(
  arrangeGrob(plist[[1]],
              do.call(arrangeGrob, c(plist[2:4], ncol=3, list(widths=c(1,1,1)))),
          nrow=2, heights=c(5,3)), legend, nrow=1, widths=c(8,1))

ggsave(filename = "multi-plot", plot = mp, path = "../images/", device = "pdf", units = "in", width = 12, height = 10)

rm(plist, g, mp, legend)
```





