---
title: "Exploring MPA and ICPP data"
output: html_notebook
author: Mark Ruddy 
date: 2017-04-20
---


## Task summary

The project aims to look at projected climate warming in Marine Protected Areas. It uses climate models ([CMIP5](http://cmip-pcmdi.llnl.gov/cmip5/)) run for two emission scenarios (Representative Concentration Pathways - [RCPs](https://www.skepticalscience.com/rcp.php?t=1)) - 4.5degC and 8.5degC - developed as part of Intergovernmental Panel on Climate Change's (IPCC) Fifth Assessment Report ([AR5](http://www.ipcc-data.org/sim/gcm_monthly/AR5/)).

As part of the study, the temperature data need to be modified to exclude land-based temperature measures. Additionally, adequate maps of the data need to be created.

This notebook will provide:

1. A section of code that removes the land-based temperatures from the ICPP surface temperature rasters.
2. Plots of SST with land surface removed (white or black) and a pleasing SST key.

## Study setup

```{r}
# Packages
library(dplyr) # Data wrangling and exploration
library(raster) # Dealing with rasters
library(ncdf4) # Required - climate data in NetCDF format
```

Datasets

```{r}
map.extent <- c(-180,180,-90,90) # Reset for layer, orginal orientation was 0-360 longitude
mean85 <- raster("../trend_yearmean_ensemble_tos_RCP85.nc")
extent(mean85) <- map.extent

mpas <- read

plot(mean85)
```


```{r}
source("../revrotate..R")
mean85 <- revrotate(mean85)
plot(mean85)

```















