---
title: "MPAs and climate change: regional sea-surface temperature means"
output:
  html_document:
    theme: readable
author: Mark Ruddy
date: 2017-04-27
---

Task is to find mean temperature values for latitudinal regions of the globe.


Packages.
```{r}
library(raster)
library(ncdf4)
library(rgdal)
library(dplyr)
library(tibble)
```

## Raster summary statistics

### Read in datasets

Read in Marine Protected Areas
```{r get-mpas}
mpas.in <- read.csv("../nn_extractedA2c.csv", header = TRUE)

mpas.xy <- mpas.in[,2:3]

rm(mpas.in)
```

Read in CMIP5 RCP4.5 max
```{r get-cmip}

cmip <- "../trend_yearmax_ensemble_RCP45.nc"
cmip <- raster(cmip) # Read in raster
map.extent <- c(-180, 180, -90, 90) # Reset for layer, orginal orientation was 0-360 longitude
extent(cmip) <- map.extent # Change extent

source("../revrotate.R") # Rotate raster layer to align with MPA points
cmip <- revrotate(cmip)
extent(cmip) <- map.extent

rm("map.extent")
```


Read in [Natural Earth](http://www.naturalearthdata.com/) 1:110m world vector, then convert to simple raster.

```{r get-land}

# Natural Earth 110m data. Comes as zip file 
url <- "http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/110m/physical/ne_110m_land.zip" # web location

localDir <- "../data"

if (!file.exists(localDir)) {
  dir.create(localDir)
}

dwnld <- paste(localDir, basename(url), sep = "/") # zip file destination

if (!file.exists(dwnld)) {
  download.file(url = url, destfile = dwnld, method = "auto") # download from web
  unzip(dwnld, exdir = localDir) # unzip
}

land <- readOGR(dsn = localDir) # Read in shapefile using `rgdal`. No need to specify shapefile name as there's only one in the directory we created above.

rm(url, localDir, dwnld) # clean up workspace
```

Remove land-surface temperatures.
```{r cmip-sst-only}
## Mask CMIP5 with land surface ##
# "cmip" is CMIP5 netCDF read in by raster
# "land" is vector polygon land surface dataset
cmip.mask <- mask(x = cmip, mask = land, inverse = TRUE)

## Get MPA cells ##
# "mpas.xy" are the MPA longitude and latitude coordinate data
mpas <- SpatialPoints(mpas.xy) # Make MPAs into SpatialPoints class

mpa.cells <- mask(x = cmip, mask = mpas)

## Merge land masked CMIP5 and MPA cells ##
cmip.sst <- merge(cmip.mask, mpa.cells)

rm(cmip.mask, mpa.cells)
```

## Find mean temperatures.

Define latitudinal zones. We're just setting Northern hemisphere limits. Mirror image southern hemisphere limits will be derived within the function to calculate zonal means.
`raster::extract` will be used, which selects cells by their centroid from `raster::extents`. This means we shouldn't have duplicated cells contributing to different zones.

```{r define-zones}
# Find global means

# Find regional means. 
# Latitudinal zone extents

global.lims <- c(0, 90)

## Tropical
trop.lims <- c(0, 23.5)

## Sub-tropical
subtrop.lims <- c(23.5, 40) # N and S subtropical zone limits.

## Temperate
temp.lims <- c(40, 66.5) # N and S temperate zone limits.

## Polar
polar.lims <- c(66.5, 90) # N and S polar zone limits.

## All extents
zone.lims <- list(global.lims, trop.lims, subtrop.lims, temp.lims, polar.lims)
```

Make data frame to hold zone names, limits and results.
```{r zones-df}
# Initialise data frame for outputs. A tibble.
zones <- c("global", "tropical", "sub-tropical", "temperate", "polar")

zone.means <- data_frame(
  zone = zones, 
  limits = zone.lims,
  means = numeric(length = length(zones)), # Mean temperature
  count = numeric(length = length(zones)) # !is.na cell count for cross-checking data capture
)

rm(list = ls(pattern = ".*ext$|.*lims$"))
```



Calculate zonal mean temperature values.
```{r }
# Function to get raster means within extents
# raster = raster of values
# zones = df of zones, limits and empty values column for mean

source("../functions/multi.extent.means.R") # Applying raster::extract over zone dataframe

zone.means.sst <- multi.extent.means(cmip.sst, zone.means)

zone.means.sst # Sea temperatures
# zone.means.all # Sea and land temperatures
```

Let's check the cells captured by these limits to make sure there's no double counting. First, we'll find the number of non-NA cells in the SST raster globally, and then compare this number against the total of non-NA cells from recombining extracts made from splitting the original raster into zones. If the totals match then the latitudinal limits used develop the regional models don't overlap - cells aren't used twice. We would expect this not to be the case as `raster::extract` selects cells based on the proximity of their centroids within extents used to define sample areas.

!is.na cells in SST raster globally
```{r}
# !is.na cells in SST raster
length(getValues(cmip.sst))-summary(cmip.sst)[6]
```

Recombined !is.na cells
```{r}
sum(zone.means.sst$count[-1])
```

Cell totals from original SST raster and recombined zonal SST extracts both match. 





