---
title: "MPAs and climate change: land surface temperature removal"
output: html_notebook
author: Mark Ruddy
date: 2017-04-22
---

## Task summary

The project aims to look at projected climate warming in Marine Protected Areas. It uses climate models ([CMIP5](http://cmip-pcmdi.llnl.gov/cmip5/)) run for two emission scenarios (Representative Concentration Pathways - [RCPs](https://www.skepticalscience.com/rcp.php?t=1)) - 4.5degC and 8.5degC - developed as part of Intergovernmental Panel on Climate Change's (IPCC) Fifth Assessment Report ([AR5](http://www.ipcc-data.org/sim/gcm_monthly/AR5/)).

As part of the study, the temperature data need to be modified to exclude land-based temperature measures. Additionally, adequate maps of the data need to be created.

This notebook will provide code that removes the land surface temperatures from the CMIP5 surface temperature rasters.

## Setup

```{r}
# Packages
library(dplyr) # Data wrangling and exploration
library(tidyr) # Data wrangling
library(raster) # Dealing with rasters
library(ncdf4) # Required - viewing and working with climate data in netCDF format
```


## Remove land-based temperatures

### Inspect CMIP5 RCP8.5 data 

Inspect structure and values using `ncdf4` package.

```{r}
nc <- nc_open(filename = name85) # Open connection to netCDF file
print(nc) # View dataset structure
nc_close(nc) # Close connection
rm("nc")
```


Inspect *Longitude and Latitude* grid.
```{r}
nc <- nc_open(filename = name85)

lon <- ncvar_get(nc = nc, varid = "lon") # Longitude named "lon" in netCDF

lat <- ncvar_get(nc = nc, varid = "lat") # Latitude named "lat" in netCDF

dim(lon); dim(lat)

head(lon); head(lat)

nc_close(nc) # Close connection

rm("lon", "lat", "nc")
```
Data is on a 1 \* 1 degree grid.

Inspect *Time* variable.
```{r}
nc <- nc_open(filename = name85)

t <- ncvar_get(nc = nc, varid = "time") # Time named "time" in netCDF

tunits <- ncatt_get(nc = nc, varid = "time", attname = "units")

dim(t)

tunits

nc_close(nc) # Close connection

rm("t", "tunits", "nc")
```
Only one time slice in the dataset.


Inspect *Temperature* variable.
```{r}
nc <- nc_open(filename = name85) # Open connection to netCDF file

print(nc) # View dataset structure

dname <- "tos" # name of variable found from `print(nc)`

tmp.df <- as_data_frame(ncvar_get(nc = nc, varid = dname)) # Get temperature (tmp) data 

colnames(tmp.df) <- 1:180

glimpse(tmp.df) # Peek at data

nc_close(nc)  # Close connection

rm("dname", "nc")
```
360 \* 180 dataset (1 \* 1 degree grid) of temperature trend values (? degree C / decade).

Plot from temperature dataframe
```{r}
# library(ggplot2) # Graphics
# library(viridis) # viridis colour scales for perceptually uniform palettes. See https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html


ggplot(mean85) + 
  geom_raster(aes(fill = value)) +
  scale_fill_viridis(direction = -1, option = "plasma", breaks = seq(0.02, 0.16, 0.02)) +
  geom_point(data = mpas, mapping = aes(x = Centroid_Longitude, y = Centroid_Latitude), size = 0.1, colour = "black") + 
  guides(fill = guide_colorbar(title = expression(paste(degree, C,"/")), barheight = 16, raster = TRUE)) +
  ggtitle("RCP 8.5 warming rate for mean SST") + 
  xlab("Longitude") +
  ylab("Latitude") +
  scale_y_continuous(breaks = seq(-80, 80, 40), expand = c(0,0)) +
  scale_x_continuous(breaks = seq(-180, 180, 60), expand = c(0,0))
```



```{r}
nc <- nc_open(filename = name85) # Open connection to netCDF file


nc_close(nc)  # Close connection
rm("nc")
```

### Obtain world land surface raster data

Read in [Natural Earth](http://www.naturalearthdata.com/) 1:110m world vector, then convert to simple raster.

```{r}
# Natural Earth 110m data. Comes as zip file 
url <- "http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/110m/physical/ne_110m_land.zip" # web location

localDir <- "../data"

if (!file.exists(localDir)) {
  dir.create(localDir)
}

dwnld <- paste(localDir, basename(url), sep = "/") # zip file destination

if (!file.exists(dwnld)) {
  download.file(url = url, destfile = dwnld, method = "auto") # download from web
  unzip(dwnld, exdir = localDir) # unzip
}

list.files(localDir) # check unzip

rm("url", "localDir", "dwnld") # clean up workspace
```



Raster calculation
<!-- Check out: -->
<!-- http://neondataskills.org/R/Raster-Data-In-R/ -->



